---
title: "Darwin Core update"
author: "Michael Maniscalco"
date: '2022-07-20'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# List of packages for session
.packages = c("googlesheets4", "readxl", "lubridate", "readr", "stringr", "janitor", "dplyr")

# Installing CRAN packages (if not already installed)
.inst <- .packages %in% installed.packages()
if(length(.packages[!.inst]) > 0) install.packages(.packages[!.inst])

# Loading packages into session 
lapply(.packages, require, character.only = TRUE)
sessionInfo()
```

## This script will load remote data of darwin core format historic data (currently in EDI database) along with new data stored on Google drive and put it together into new darwin core files

------------------------------------------------------------------------

### Change name and file path for output files from the example_name\_\*.csv below

```{r out_name}
occurence_out_name<- "20220720occurence_example_name.csv"
event_out_name<- "20220720event_example_name.csv"
extended_out_name<- "20220720extended_example_name.csv"
```

### Read in data

-   could change paths to EDI links rather than gdrive

```{r read }

googlesheets4::gs4_auth(path = '.secrets/enduring-coil-349821-3832e38b5b43.json')

df_event <- googlesheets4::read_sheet("https://docs.google.com/spreadsheets/d/1BaCEGePEYx9wbduccHytb9uRVbz3447GEDCE8Hpm0I8/edit#gid=1448123747",
                                      sheet="Event")

df_phyto <- googlesheets4::read_sheet("https://docs.google.com/spreadsheets/d/1BaCEGePEYx9wbduccHytb9uRVbz3447GEDCE8Hpm0I8/edit#gid=1550025152",
                                      sheet="Occurrence")

# ## env
df_envir <- googlesheets4::read_sheet("https://docs.google.com/spreadsheets/d/1BaCEGePEYx9wbduccHytb9uRVbz3447GEDCE8Hpm0I8/edit#gid=1448123747",
                                      sheet="Extended")
```

```{r format}
df_phyto_wide <- df_phyto %>%
  dplyr::select(-c(occurrenceID:kingdom)) %>%
  tidyr::pivot_wider(.,names_from = organismName,values_from = organismQuantity)

df_envir_wide <- df_envir %>%
  dplyr::select(-c(measurementID:measurementRemarks)) %>%
  tidyr::pivot_wider(.,names_from = measurementType,values_from = measurementValue)

historical_wide <- dplyr::left_join(df_event,df_envir_wide) %>%
  dplyr::left_join(.,df_phyto_wide) %>%
  dplyr::relocate(monitor_names, .after="eventRemarks")%>%
  dplyr::relocate(descending_transparency:ascending_transparency, .before="transparency_depth_mean") %>%
  dplyr::mutate(across(c(air_temp:water_temp, wind_speed_knots,
                  tide_height:bod, pH, bottom_depth_m,
                   wind_direction), ~as.numeric(.x))) %>%
  dplyr::mutate(time_high_tide=as.POSIXct(time_high_tide,
                                   format="%Y-%m-%d %H:%M:%S",
                                   tz ="UTC"),
         time_low_tide=as.POSIXct(time_low_tide,
                                  format="%Y-%m-%d %H:%M:%S",
                                  tz ="UTC")) 
tax_table <- readr::read_csv("source_data/tax_table.csv")

df_new <- googlesheets4::read_sheet("https://docs.google.com/spreadsheets/d/1JGXLAvG_U9dWn7Ya_NorUU3mcgpi7aX2uanszfhXP8c/edit?sharingaction=ownershiptransfer#gid=0",
                                    sheet = "processed") %>%
  dplyr::mutate(across(c(water_temp:air_temp,rainfall_mm,
                         wind_speed_knots,
                         ascending_transparency:nutrient_vial_id, 
                         wind_direction),
                       ~as.numeric(.x)))%>%
  dplyr::select(-c(Copepods, Other_zooplankton, Total_zooplankton, sampling_method)) %>%
  dplyr::mutate(maximumDepthInMeters=ifelse(is.na(maximumDepthInMeters)==T,
                                            1,
                                            maximumDepthInMeters),
                minimumDepthInMeters=ifelse(is.na(minimumDepthInMeters)==T,
                                            0.5,
                                            minimumDepthInMeters)) 
### wide
df_full <- dplyr::full_join(historical_wide,df_new)
rm(df_new,historical_wide, df_event, df_phyto, df_envir, df_phyto_wide ,df_envir_wide)
```

### Darwin core Occurrence

```{r Occur}
df_full_occurrence <- df_full %>%
  dplyr::rowwise() %>%
  tidyr::pivot_longer(., cols=(Alexandrium:Total_phytoplankton),
                      names_to = "organismName",
                      values_to = "organismQuantity") %>%
  dplyr::left_join(.,tax_table, by="organismName") %>%
  dplyr::mutate(occurrenceID=paste(strftime(eventDate, "%Y-%m-%dT%H:%M:%S%Z", tz = "Zulu"), organismName, sep="_")) %>%
  dplyr::mutate(occurrenceStatus= ifelse(organismQuantity>0, "present", "absent")) %>%
  dplyr::select(id, basisOfRecord, organismName, organismQuantity, organismQuantityType,
                occurrenceID, occurrenceStatus, scientificName, scientificNameID, taxonID, kingdom) %>%
  dplyr::filter_at(vars(occurrenceStatus), all_vars(!is.na(.))) 

### add new file name/location
readr::write_csv(df_full_occurrence,paste0(occurence_out_name))
```

### Darwin core Events

```{r event}
df_full_events <- df_full %>%
  dplyr::select(id, eventID,	eventDate, geodeticDatum,	countryCode, locationID, 
         eventRemarks,	decimalLatitude,	decimalLongitude,	minimumDepthInMeters,
         maximumDepthInMeters,	coordinateUncertaintyInMeters)%>%
  dplyr::arrange(eventDate)

### add new file name/location
readr::write_csv(df_full_events,paste0(event_out_name))
```

### Make ExtendedMeasurementOrFact

```{r extended}
### Darwin core Extended
extended_base <- readr::read_csv("source_data/columns_units.csv") %>%
  dplyr::filter(darwin_file=="all"|darwin_file=="ExtendedMeasurementOrFact")

phyto_extended <- df_full %>%
  dplyr::select(-eventRemarks, -nutrient_vial_id, 
         -c(Alexandrium:Other_phytoplankton)) %>% 
  dplyr::mutate(across(air_temp:transparency_depth_mean, ~as.character(.x))) %>%
  tidyr::pivot_longer(., 
               cols=(time_low_tide:pH),
               names_to = "measurementType",
               values_to = "measurementValue",
               values_transform = list(measurementValue = as.character)) %>%
  dplyr::mutate(measurementValue=ifelse(measurementType=="rainfall_mm" & measurementValue=="0",
                                 NA, 
                                 measurementValue)) %>%
  dplyr::filter(!is.na(measurementValue)) %>%
  dplyr::left_join(., extended_base, by= "measurementType") %>%
  dplyr::select(.,id, measurementType,	measurementValue,	measurementID,
         measurementTypeID,	measurementUnit,
         measurementUnitID, # is this one optional?
         measurementRemarks) %>%
  arrange(id)

### add new file name/location here 
readr::write_csv(phyto_extended, paste0(extended_out_name))
```
