---
title: "MDIBL phyto data into Darwin Core Archive format"
author: "Michael Maniscalco"
date: '2022-06-29'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readxl)
library(tidyverse)
library(lubridate)
library(janitor)
```

### Load tow and filtered sample data sets

```{r load, include=F}
df_phyto <- googlesheets4::read_sheet("https://docs.google.com/spreadsheets/d/1gj44rb7tz8_Y43Io3aIVJaJfgN-Gr9mOGEokjDZpu5E/edit#gid=902631445") %>%
  arrange(eventDate) %>%
  mutate(maximumDepthInMeters=ifelse(is.na(maximumDepthInMeters)==T,
         1,
         maximumDepthInMeters),
         minimumDepthInMeters=ifelse(is.na(minimumDepthInMeters)==T,
         0.5,
         minimumDepthInMeters)) 

df_tow <- googlesheets4::read_sheet("https://docs.google.com/spreadsheets/d/1bz9ZMGaARpJCBnHZB632ltLvTyscNlLQsaqxae6SWqQ/edit#gid=0") %>%
  arrange(eventDate)

df_full <- full_join(df_phyto, df_tow)
```

### Create occurrence table and save as csv
* Pivot phytoplankton count data to long 
* Subset entries when only a subset of target species were counted and convert 0 counts to NA for other taxa and remove from long form data
* Add additional Darwin core fields
* Select columns to keep in desired order
* Occurrence file

```{r occur}
tax_table <- read_csv("source_data/tax_table.csv") %>%
  select(-names_originally_in_access_db,-names_in_anecdata)

phyto_occurrence <- df_phyto %>%
  rowwise() %>%
  pivot_longer(., cols=(Alexandrium:Other_phytoplankton),
                 names_to = "organismName",
               values_to = "organismQuantity",) %>%
  left_join(.,tax_table, by="organismName") %>%
  add_column(basisOfRecord="HumanObservation", .after="id") %>%
  add_column(organismQuantityType= "Cells per L", .after="organismQuantity") %>% 
  mutate(organismQuantity=ifelse(is.na(organismQuantity)==T, 0, organismQuantity)) %>%
  mutate(occurrenceID=paste(strftime(eventDate , "%Y-%m-%dT%H:%M:%S%Z", tz = "Zulu"), 
                            organismName, sep="_")) %>%
  mutate(occurrenceStatus= ifelse(organismQuantity>0, "present", "absent")) %>%
  filter(only_dinophysis_counted==TRUE & str_detect(organismName,"Dinophysis")==T |only_alexandrium_counted==TRUE & str_detect(organismName,"Alexandrium")==T |only_PN_counted==TRUE & str_detect(organismName,"Pseudo_nit" )==T) %>%
  filter(organismName != "Dinophysis_acuminata") %>%
  filter(organismName != "Dinophysis_norvegica" )

phyto_occurrence_b <- df_phyto %>%
  rowwise() %>%
  pivot_longer(., cols=(Alexandrium:Other_phytoplankton),
                 names_to = "organismName",
               values_to = "organismQuantity",) %>%
  left_join(.,tax_table, by="organismName") %>%
  add_column(basisOfRecord="HumanObservation", .after="id") %>%
  add_column(organismQuantityType= "Cells per L", .after="organismQuantity") %>% 
  mutate(organismQuantity=ifelse(is.na(organismQuantity)==T, 0, organismQuantity)) %>%
  mutate(occurrenceID=paste(strftime(eventDate , "%Y-%m-%dT%H:%M:%S%Z", tz = "Zulu"), organismName, sep="_")) %>%
  mutate(occurrenceStatus= ifelse(organismQuantity>0, "present", "absent")) %>%
  filter(is.na(only_dinophysis_counted)==T &
           is.na(only_alexandrium_counted)==T & 
           is.na(only_PN_counted)==T ) 

phyto_occurrence <- bind_rows(phyto_occurrence, phyto_occurrence_b) %>%
  select(id, basisOfRecord,organismName, organismQuantity, organismQuantityType,
         occurrenceID, occurrenceStatus, scientificName, scientificNameID, taxonID, kingdom) %>%
  mutate(occurrenceStatus=ifelse(occurrenceStatus=="absent" &
                                   !str_detect(organismName,"Dinophysis|Pseudo_n|Alexandrium|Prorocentrum|Margalefidinium|Karenia"),
                                 NA,
                                 occurrenceStatus)) %>%
  filter_at(vars(occurrenceStatus), all_vars(!is.na(.)))


tow_occurrence <- df_tow %>%
    pivot_longer(., cols=(Alexandrium:Other_phytoplankton),
               names_to = "organismName",
               values_to = "organismQuantity") %>%
  filter(organismQuantity>0) %>%
  left_join(.,tax_table, by="organismName") %>%
  add_column(basisOfRecord="HumanObservation", .after="id") %>%
  add_column(organismQuantityType= "Relative Abundance", .after="organismQuantity") %>% 
  mutate(occurrenceID=paste(eventDate, organismName, sep="_")) %>%
  add_column(occurrenceStatus="present") %>%
  select(id, basisOfRecord,organismName, organismQuantity, organismQuantityType,
         occurrenceID, occurrenceStatus, scientificName, scientificNameID, taxonID, kingdom)

phyto_occurrence <- bind_rows(tow_occurrence,phyto_occurrence)

write_csv(phyto_occurrence,"2022_mdibl_data_package/data_objects/occurrence.csv")

googlesheets4::gs4_auth(path = '.secrets/enduring-coil-349821-3832e38b5b43.json')
googlesheets4::write_sheet(phyto_occurrence, ss="https://docs.google.com/spreadsheets/d/1BaCEGePEYx9wbduccHytb9uRVbz3447GEDCE8Hpm0I8/edit#gid=1550025152",sheet = "Occurrence")

rm(phyto_occurrence, tow_occurrence, phyto_occurrence_b)
```

### Create event file

```{r event}
phyto_events <- df_full %>%
  select(id, eventID,	eventDate, geodeticDatum,	countryCode, locationID, 
         eventRemarks,	decimalLatitude,	decimalLongitude,	minimumDepthInMeters,
         maximumDepthInMeters,	coordinateUncertaintyInMeters)%>%
  arrange(eventDate)
write_csv(phyto_events,"2022_mdibl_data_package/data_objects/event.csv")

googlesheets4::gs4_auth(path = '.secrets/enduring-coil-349821-3832e38b5b43.json')
googlesheets4::write_sheet(phyto_events, ss="https://docs.google.com/spreadsheets/d/1BaCEGePEYx9wbduccHytb9uRVbz3447GEDCE8Hpm0I8/edit#gid=0",sheet = "Event")
rm(phyto_events)
```

### Create ExtendedMeasurementOrFact table

```{r extended}
extended_base <- read_csv("source_data/columns_units.csv") %>%
  filter(darwin_file=="all"|darwin_file=="ExtendedMeasurementOrFact")

phyto_extended <- df_full %>%
  select(-eventRemarks, -nutrient_vial_id, 
         -c(Alexandrium:Other_phytoplankton)) %>% 
  mutate(across(air_temp:transparency_depth_mean, ~as.character(.x))) %>%
  relocate(nitrate_plus_nitrite:silicate, .after=pH) %>%
  pivot_longer(., 
               cols=(time_low_tide:silicate ),
               names_to = "measurementType",
               values_to = "measurementValue",
               values_transform = list(measurementValue = as.character)) %>%
  mutate(measurementValue=ifelse(measurementType=="rainfall_mm" & measurementValue=="0",
                                 NA, 
                                 measurementValue)) %>%
  filter(!is.na(measurementValue)) %>%
  left_join(., extended_base, by= "measurementType") %>%
  select(.,id, measurementType,	measurementValue,	measurementID,
         measurementTypeID,	measurementUnit,
         measurementUnitID, # is this one optional?
         measurementRemarks) %>%
  arrange(id)

write_csv(phyto_extended,"2022_mdibl_data_package/data_objects/ExtendedMeasurementOrFact.csv")

googlesheets4::gs4_auth(path = '.secrets/enduring-coil-349821-3832e38b5b43.json')
googlesheets4::write_sheet(phyto_extended, ss="https://docs.google.com/spreadsheets/d/1BaCEGePEYx9wbduccHytb9uRVbz3447GEDCE8Hpm0I8/edit#gid=1448123747", sheet = "Extended")
rm(phyto_extended, extended_base)
```

### Write temp phyto for Rshiny visualization

```{r }
ss <- googlesheets4::gs4_get("https://docs.google.com/spreadsheets/d/1JGXLAvG_U9dWn7Ya_NorUU3mcgpi7aX2uanszfhXP8c/edit#gid=641716790")


phyto_occurrence <- df_phyto %>%
  rowwise() %>%
  pivot_longer(., cols=(Alexandrium:Other_phytoplankton),
                 names_to = "organismName",
               values_to = "organismQuantity",) %>%
  left_join(.,tax_table, by="organismName") %>%
  add_column(basisOfRecord="HumanObservation", .after="id") %>%
  add_column(organismQuantityType= "Cells per L", .after="organismQuantity") %>% 
  mutate(organismQuantity=ifelse(is.na(organismQuantity)==T, 0, organismQuantity)) %>%
  mutate(occurrenceID=paste(strftime(eventDate , "%Y-%m-%dT%H:%M:%S%Z", tz = "Zulu"), 
                            organismName, sep="_")) %>%
  mutate(occurrenceStatus= ifelse(organismQuantity>0, "present", "absent")) %>%
  filter(only_dinophysis_counted==TRUE & str_detect(organismName,"Dinophysis")==T |only_alexandrium_counted==TRUE & str_detect(organismName,"Alexandrium")==T |only_PN_counted==TRUE & str_detect(organismName,"Pseudo_nit" )==T) %>%
  filter(organismName != "Dinophysis_acuminata") %>%
  filter(organismName != "Dinophysis_norvegica" )

phyto_occurrence_b <- df_phyto %>%
  rowwise() %>%
  pivot_longer(., cols=(Alexandrium:Other_phytoplankton),
                 names_to = "organismName",
               values_to = "organismQuantity",) %>%
  left_join(.,tax_table, by="organismName") %>%
  add_column(basisOfRecord="HumanObservation", .after="id") %>%
  add_column(organismQuantityType= "Cells per L", .after="organismQuantity") %>% 
  mutate(organismQuantity=ifelse(is.na(organismQuantity)==T, 0, organismQuantity)) %>%
  mutate(occurrenceID=paste(strftime(eventDate , "%Y-%m-%dT%H:%M:%S%Z", tz = "Zulu"), organismName, sep="_")) %>%
  mutate(occurrenceStatus= ifelse(organismQuantity>0, "present", "absent")) %>%
  filter(is.na(only_dinophysis_counted)==T &
           is.na(only_alexandrium_counted)==T & 
           is.na(only_PN_counted)==T ) 

phyto_occurrence <- bind_rows(phyto_occurrence, phyto_occurrence_b) %>%
  mutate(occurrenceStatus=ifelse(occurrenceStatus=="absent" &
                               !str_detect(organismName,"Dinophysis|Pseudo_n|Alexandrium|Prorocentrum|Margalefidinium|Karenia"),
                                 NA,
                                 occurrenceStatus)) %>%
  filter_at(vars(occurrenceStatus), all_vars(!is.na(.)))


tow_occurrence <- df_tow %>%
  pivot_longer(., cols=(Alexandrium:Other_phytoplankton),
               names_to = "organismName",
               values_to = "organismQuantity") %>%
  filter(organismQuantity>0) %>%
  left_join(.,tax_table, by="organismName") %>%
  add_column(basisOfRecord="HumanObservation", .after="id") %>%
  add_column(organismQuantityType= "Relative Abundance", .after="organismQuantity") %>% 
  mutate(occurrenceID=paste(eventDate, organismName, sep="_")) %>%
  add_column(occurrenceStatus="present") 

phyto_occurrence <- bind_rows(tow_occurrence,phyto_occurrence)%>%
  filter(organismQuantityType=="Cells per L") %>%
  dplyr::mutate(day=strftime(eventDate, "%Y-%m-%d", tz = "America/New_York")) %>%
  dplyr::select(day, locationID,  organismName, organismQuantity)

googlesheets4::gs4_auth(path = '.secrets/enduring-coil-349821-3832e38b5b43.json')
googlesheets4::write_sheet(ss, sheet = "occurrence_temp", data=phyto_occurrence)
rm(tax_table, phyto_occurrence,phyto_occurrence_b, tow_occurrence)
```

### Write extended temporary

```{r}
phyto_extended <- df_full %>%
  dplyr::relocate(.,c(salinity_ppt, DOavg_ppm,bod, transparency_depth_mean, 
                      orthophosphate, silicate, ammonium, nitrate_plus_nitrite),
                  .after="water_temp") %>%
  dplyr::select(-eventRemarks, -nutrient_vial_id, 
         -c(Alexandrium:only_alexandrium_counted)) %>% 
  tidyr::pivot_longer(., 
               cols=(water_temp:nitrate_plus_nitrite),
               names_to = "measurementType",
               values_to = "measurementValue",
               values_transform = list(measurementValue = as.character)) %>%
  dplyr::filter(!is.na(measurementValue)) %>%
  dplyr::mutate(day=strftime(eventDate, "%Y-%m-%d", tz = "America/New_York"), 
                measurementValue=as.numeric(measurementValue)) %>%
  dplyr::select(.,day, locationID,	measurementType, 
         measurementValue) %>%
  dplyr::arrange(day)

googlesheets4::gs4_auth(path = '.secrets/enduring-coil-349821-3832e38b5b43.json')
googlesheets4::write_sheet(ss, sheet = "extended_temp", data=phyto_extended)
rm(phyto_extended, ss, df_full, df_phyto, df_tow)
```

```{r}
```
