---
title: "Anecadata phyto clean"
author: "Michael Maniscalco"
date: '2022-06-27'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readxl)
library(tidyverse)
library(lubridate)
library(janitor)
```

### Load Anecdata phyto download
* Remove unwanted fields
* Separate "hotspot" into locationID and site name
* Rename several columns into a more machine/human readable format
* Concatenate various comments fields into one Column

```{r load, include=FALSE}
anecdata_db <- read_csv("source_data/maine_phytoplankton_monitoring-2022-06-13_03.21.46.csv") %>%
  filter(!str_detect(username,"MarkWhiting")) %>%
  select(-license, 
         -license_url, 
         -geoprivacy, 
         -event, 
         -c(country:postcode),
         -c(observation_notes:Abundance), 
         -`Nutrient_bath_temperature_(C)`,
         -c("Phytoplankton_(Dominant_Species_1)":"Date_e_mailed_to_DMR"),
         -c("Image 1":"Image 5"),
         -Microplastics_Sampling_Type,
         -c("Blue__Round":"Microplastics_Notes"),
         -c("Frequency_(Dominant_Species_1)":"Frequency_(Dominant_Species_2)"),
         -Zooplankton_Frequency,
         -username,
         -id,
         -row_id) %>%
  separate(., col = hotspot, 
           into = c("locationID",
                    "site"),
           sep = " ",
           remove = T,
           extra = "merge") %>%
  rename(
        decimalLatitude="lat",
        decimalLongitude="lng",
        # "site",
        eventDate ="observed",
        water_temp="Water_Temperature_(_C)_(C)",
        air_temp="Air_temperature_(_C)_(°C)",
        transparency_depth_mean="Transparency_(m)",
        salinity_ppt="Salinity_(ppt)",
        DOavg_ppm="Dissolved_Oxygen",
        water_current="Current_(relative)",
        nutrient_vial_id="Nutrient_Vial_Number",
        Pseudo_nitzschia_large="__of_LARGE_Pseudo_nitzschia",
        Pseudo_nitzschia_small="__of_SMALL_Pseudo_nitzschia",
        Dinophysis_acuminata="__of_Dinophysis_acuminata",     
        Alexandrium="__of_Alexandrium",   
        Dinophysis_norvegica="__of_Dinophysis_norvegica", 
        Dinophysis_spp="__of_Dinophysis_spp__(unknown)", 
        Prorocentrum_lima="__of_Prorocentrum_lima", 
        Margalefidinium_polykrikoides="__of_Margalefidinium_p_", 
        Karenia="__of_Karenia", 
        too_numerous_to_count_grid_num="PS_TNTC_reached_at_cell__", 
        lab_comments="Lab_Comments", 
        monitor_names="Collected_By", 
        rain_inches="Inches_of_Rain_(in)", 
        tide_stage="Tide_in___out",                 
        time_low_tide="Time_of_Low_Tide", 
        time_high_tide="Time_of_High_Tide", 
        weather="Weather", 
        water_surface="Water_Surface",                 
        wind_direction="Wind_Direction", 
        sampling_method="Phytoplanton_Sample_Method", 
        field_comments= "Field_Notes", 
        descending_transparency="Descending_Transparency_(m)", 
        ascending_transparency="Ascending_Transparency_(m)", 
        bottom_depth_m="Water_Depth_(m)",               
        water_filtered_L="Amount_of_Water_Filtered_(L)",  
        wind_description= "Wind_Speed", 
        grids_counted = "__Grid_Cells_Counted", 
        Copepods= "__of_Copepods",
        wind_speed_knots="Wind_Speed_(mph)") %>%
  unite(., eventRemarks, lab_comments,Comments,field_comments, sep = "|", remove = TRUE)
```

### Normalize PN counts to proportion of grids counted
* In many cases when counts of PN reached 500 cells the grid number was recorded and then only other cells were counted for the remainder of the slide. The number of grids counted needs to be used to extrapolate out to the total number of cells that would have been counted if the standard 200 grids were counted

```{r drop}
anecdata_db <- anecdata_db %>%
  mutate(Pseudo_nitzschia_large=Pseudo_nitzschia_large*200/grids_counted,
         Pseudo_nitzschia_small=Pseudo_nitzschia_small*200/grids_counted)
```

### Tide times unto UTC datetime for easier comparison with eventDate

```{r drop}
anecdata_db <- anecdata_db %>%
  mutate(time_low_tide = as.POSIXct(paste(
    lubridate::as_date(as.POSIXct(eventDate, tz ="America/New_York")),
    time_low_tide), format="%Y-%m-%d %H:%M:%S", tz ="America/New_York"),
    time_high_tide = as.POSIXct(paste(
    lubridate::as_date(as.POSIXct(eventDate, tz ="America/New_York")),
    time_high_tide), format="%Y-%m-%d %H:%M:%S", tz ="America/New_York")
         )
```

### Fix any site name and locationID inconsistencies

```{r siteIDs}
anecdata_db <- anecdata_db %>%
  select(-too_numerous_to_count_grid_num) %>% # the too numerous to count was no longer used after migration of database to anecdata
  mutate(site=str_replace(site, "MDIBL dock|mdibl dock|.*MDIBL.*","MDIBL Dock"))%>%
  mutate(site= str_replace(site, "eastern bay channel", "Eastern Bay Channel"),
         site= str_replace(site, "Surry - Town Landing", "Surry Town Pier"),
         site= str_replace(site, "Stave", "Stave Island"),
         site= str_replace(site, "Lamoine Shore", "Lamoine Beach"),
         site= str_replace(site, "Brooksville Town Landing", "Brooksville Town Dock"))

site_names_corrected <- read_csv("source_data/site_details.csv")

anecdata_db <- left_join(select(anecdata_db,
                                -c(locationID, decimalLatitude, decimalLongitude)),
                      site_names_corrected, by="site") %>%
  select(-locationID) %>%
  rename(locationID=suggested_locationID) %>%
  relocate(locationID:coordinateUncertaintyInMeters, .after=site)
rm(site_names_corrected)
```

### Convert rain from inches to millimeters for standard metric units

```{r rain}
anecdata_db <- anecdata_db %>%
  mutate(rainfall_mm=rain_inches*25.4) %>%
  select(-rain_inches) 
```

### Time stage
* Shorten to just "In" or "Out"

```{r tide_stage}
anecdata_db <- anecdata_db %>%
  mutate(tide_stage=str_replace(tide_stage, pattern = "In \\(tide is rising\\)", replacement = "In"),
         tide_stage=str_replace(tide_stage, pattern="Out \\(tide is falling\\)", replacement = "Out"))
```

### Convert wind direction from text strings to degrees
* Converting the single letter abbreviations into words and vis-versa with grep/stringr is prone to more problems/inconsistencies

```{r wind_direction}
anecdata_db <- anecdata_db %>%
  mutate(wind_direction=str_replace(wind_direction, "-No Wind","999"),
         wind_direction=str_replace(wind_direction, "Northwest", "315"),
         wind_direction=str_replace(wind_direction, "Northeast", "35"),
         wind_direction=str_replace(wind_direction, "Southeast", "135"),
         wind_direction=str_replace(wind_direction, "Southwest", "225"),
         wind_direction=str_replace(wind_direction, "West", "270"),
         wind_direction=str_replace(wind_direction, "East", "90"),
         wind_direction=str_replace(wind_direction, "North", "0"),
         wind_direction=str_replace(wind_direction, "South", "90"),
         wind_direction=as.numeric(wind_direction))
```

### Add additional Darwin Core columns

```{r darwin columns}
anecdata_db <- anecdata_db %>%
  add_column(countryCode="US") %>%
  mutate(eventID= paste(locationID, 
                        strftime(eventDate , "%Y-%m-%dT%H:%M:%S%Z", tz = "Zulu"), 
                        sep="_")) %>% 
  mutate(id= paste("MDIBL-habs", locationID, 
                   strftime(eventDate , "%Y-%m-%dT%H:%M:%S%Z",tz = "Zulu"), 
                   sep="_")) %>% 
  add_column(geodeticDatum="EPSG:4326 WGS84")
```

### Convert wind speed to knots from mph

```{r drop}
anecdata_db <- anecdata_db %>%
  mutate(wind_speed_knots=wind_speed_knots*0.868976)
```

### Reformat monitor names

```{r monitors}


```

### Join with historic data of filtered samples and rearrange columns

```{r join, include=FALSE}
phyto_db <- read_csv("intermediate/MDIBL_phyto_db_filter2007_2021.csv")
phyto_full <- bind_rows(phyto_db, anecdata_db) %>%
  relocate(Dinophysis_acuminata:Margalefidinium_polykrikoides, .before="Dinophysis_spp") %>%
  relocate(Karenia, .after="Gyrosigma") %>%
  relocate(Dinophysis_norvegica, .after="Dinophysis_spp") %>% 
  relocate(wind_description, .before="water_filtered_L") %>% 
  relocate(pH:Copepods, .before="water_filtered_L") %>%
  relocate(nutrient_vial_id:Copepods, .before="sampling_method") %>%  
    relocate(Copepods, .before="Alexandrium") %>%  
  relocate(time_low_tide:time_high_tide, .after="eventDate") %>%
  select(-total_phytoplankton) %>%
  add_column(backwashed_volume= 0.015,.after = "grids_counted")

rm(anecdata_db, phyto_db)
```

### Convert counts to counts per L
* Harmonize the two columns that recorded the number of slide grids counted
* The default number was 200 grids. If no number recorded than 200 were counted.
* The default volume filtered was 10 L. If no number recorded than 10L was filtered.
* The default resuspension volume  was 0.015 L. If no number recorded than 0.015 L was volume.

```{r convert}
## counts are in cells/
total_rafter_grids <- 1000 # per slide
volume_of_rafter <- .001

phyto_full <- phyto_full %>%
  mutate(grids_counted=ifelse(is.na(grids_counted)==T, 200, grids_counted),
         grids_counted=ifelse(grids_counted==0, NA, grids_counted),
         water_filtered_L=ifelse(is.na(water_filtered_L)==T, 10, water_filtered_L)) %>%
  mutate(across(Alexandrium:Other_phytoplankton, 
                ~.x* ((1/grids_counted) *   #convert counts to counts per grid
                        (total_rafter_grids/volume_of_rafter) * # counts per grid to counts per L
                        (backwashed_volume/water_filtered_L)))) %>% # factor of concentration/dilution
  select(-BOD_date, -date_emailed_to_DMR, 
         -c("data_for_Kate", "only_PN_counted", "only_dinophysis_counted", 
            "only_alexandrium_counted", "grids_counted", "backwashed_volume",
            "water_filtered_L")) %>%
  relocate(Margalefidinium_polykrikoides, .after="Melosira")
rm(volume_of_rafter, total_rafter_grids)
```

### Write 2007-present counts per L flat file

```{r write}
write_csv(phyto_full,"source_data/phyto_db_filtered_2007_present_cells_L.csv")
googlesheets4::write_sheet(phyto_full, ss="https://docs.google.com/spreadsheets/d/1gj44rb7tz8_Y43Io3aIVJaJfgN-Gr9mOGEokjDZpu5E/edit#gid=0",sheet = "Sheet1")
head(phyto_full)
rm(phyto_full)
```

### Convert to forward compatable format

```{r }

```



```{r }

```



